---
title: "Markdown Commands"
author: "Noah W.K. Mattheis"
date: "2025-01-21"
output: 
    html_document: 
      toc: true
      highlight: tango
      theme: cosmo
    pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Ok so I can just type as I normally would for any text, no marks needed. 
Always seperate paragraphs with two carriage returns.


Here is our second paragraph.

Header size determined by number of "#" with less being larger font

# First header (largest)

## If done this way, will print all in one line
Noah Mattheis
Pharmacology MS Student
University of Vermont

## Second Level (getting smaller in size)
## need two spaces between them for lines between


### Third level

#### Fourth level
Noah Mattheis

Pharmacology MS Student

University of Vermont

Use *astericks* for *italic* and **double astericks** for **bold**

Use ^caret^ '^' for ^superscript^ and ~tilda~ '~' for ~subscript~ and ~~2 tildas~~ '~~'for ~~strikethrough~~

Use r in an in-line text:

The value of pi + 10 is `r pi + 10`.

> Use a greater than sign for indented quoted material. Looks nice Ceaser-chan! 

## For Lists

- first item
- second item
    - indented list item
    + also indented item
    
## For Links

links [linking text](website address)
Lets try for my favorite magic card art
    
  Bearer of the Heavens [Card](https://gatherer.wizards.com/pages/card/details.aspx?name=Bearer%20of%20the%20Heavens)
    
## For Tables

First Header | Second Header
------------ | -------------
Content Cell | Content Cell
Content Cell | Content Cell

## Fencing Options

Use a single back tick/ backwards quotation mark for `in line fencing` of material.

Use triple back tips for fenced block of text

```
Everything here is a plain text
even with markdown *commands*
Stays until closed with 3 back tips

Blank lines properly spaced

Very nice, Ceaser-chan
```

## Important type of fencing
#### using chunk of r code - everything between the marks will be executed as r code
#### will show code and execution of said code
#### if eval=FALSE, will not show code output, easier for sharing code
#### if eval=TRUE AND echo=FALSE, we will see output but not the code

```{r, eval=TRUE, echo=FALSE}
# r code can be used here! 

print(runif(10))
plot(runif(10), runif(10))


```

Having both be true

```{r, eval=TRUE, echo=TRUE}
# r code can be used here! 

print(runif(10))
plot(runif(10), runif(10))


```

Having eval false here

```{r, eval=FALSE}
# r code can be used here! 

print(runif(10))
plot(runif(10), runif(10))


```

## Using LaTeX for Math, o ya baby

Use a single dollar sign at the beginning and end of equation $a = b + c$ 

To insert a mathematical statement within plain text, no spaces
OR use double dollar signs, can be used with spaces, for center and separted equation 
$$ a = b + c $$
Subscripts, using '_'

'H_0" important here
$$H_0 = Z_{a + b}$$
Superscripts, using a caret '^'
$$S = cA^z$$
Combining 

$$S=cA^z_1 + z_{2 + x}$$

Fraction with variables

$$\alpha = \frac{\beta}{\delta + \gamma_x}$$

Summation signs

$$z = \sum_{i=1}^X{K}$$
Just a backslash

$$\backslash$$
backslash le gives us less than or equal to
$$\backslash \alpha \le b \backslash$$

Mixing text and equations, we need an m box, without it it will smush text together and treat it all as one variable
$$P(\mbox{Occurence of Species A}) = Z$$

Back to chunkin'
```{r}
#Here is a new chunk of code, distant from the first one in our document.

z <- 1:10
print(z)
#end of second chunk
```
