---
title: "For_Loops"
author: "Noah W.K. Mattheis"
date: "2025-02-25"
output: 
  html_document: 
    highlight: tango
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Going over Loops/For Loops/ While Loops

For Loops can be slow in R, especially when using functions like cbind in a loop \n
R is a language that doesn't need to rely on them too much \n
So why do we use them? \n
Comfort, ease of transfer to other languages, sometimes easier to read, and sometimes laziness \n


## BASIC ANATOMY OF FOR LOOP
## for (var in seq) { start of for loop
##
##  body of for loop
##
## } end of for loop

Good rule of thumb - Use i, j, or k for the var of loop \n
This way, we know if we are using these counter variables, we know we are in a loop \n
Common etiquette across other languages as well, like Java and Python \n

#### General way to set up for loops
```{r}
for (i in 1:5) { # initializes loop and will iterate 5 times
  cat("stuck in a loop", "\n") # First will print/concatenate this line
  cat(3 + 2, "\n") # then will do this arithmetic
  cat(runif(1), "\n") # then will create a rand unif number
} # Loop will run 5 times, as we specified, before it is finished
```

#### For loops can also iniate/reassign values
```{r}
z <- 0
for (i in 1:5) { # initializes loop will iterate 5 times
  z <- z + i
  print(z)
}
```

#### For loop with iterations based on vector length
```{r}
my_dogs <- c('chow', 'puggle', 'pug', 'st bernard', 'husky') # Creates vector of strings
for (i in 1:length(my_dogs)) { # i will iterate first slot of vector to end of vector
  cat("i =", i, "my_dogs[i] =", my_dogs[i], "\n")
}
```

#### Using NULL, showing length of this vector
```{r}
my_bad_dogs <- NULL
for (i in 1:length(my_bad_dogs)) {
  cat("i =", i, "my_bad_dogs[i] =", my_bad_dogs[i], "\n")
} # Code has 2 print outs, because we declared it to run at 1, but NULL is = 0, so it counts back to 0
```

#### Same as first for loop with my_dogs
```{r}
for (i in seq_along(my_dogs)) { #seq_along creates sequence following length of vector
  cat("i =", i, "my_dogs[i] =", my_dogs[i], "\n")
}
```

#### Prints out nothing, compared to 2 print outs
```{r}
for (i in seq_along(my_bad_dogs)) {
  cat("i =", i, "my_bad_dogs[i] =", my_bad_dogs[i], "\n")
}
```

#### Using seq_len for loop range and comparing to length
```{r}
zz <- 5
for (i in seq_len(zz)) {
  cat("i =", i, "my_dogs[i] =", my_dogs[i], "\n")
}

for (i in length(my_dogs)) {
  my_dogs[i] <- toupper(my_dogs[i])
  cat("i =", i, "my_dogs =", my_dogs[i], "\n")
}

my_dogs <- tolower(my_dogs)
my_dogs <- toupper(my_dogs)
my_dogs


my_dat <- runif(1)
for (i in 2:10) {
  temp <- runif(1)
  my_dat <- c(my_dat, temp)
  cat("loop number =", i, "vector_element =", my_dat[i], "\n")
}
length(my_dat)
```

## TIP: Don't write a loop if operation can be vectorized

# Just a reminder of division
```{r}
16/5
16 %% 5
16 %/% 5 # returns how many times following number can divide into first number
```

#### What if we want to only work with odd-numbered elements?
#### %% is used to return remainder after division
#### Using i %% 2==0, were saying if the remainder equals 0, i.e. is divisible by 2
#### we skip it using 'next',which takes us back to beginning of loop
```{r}
z <- 1:20
for(i in 1:length(z)) {
  if(i %% 2==0) next
  print(i)
}
```

# Same loop but now returning even numbers
```{r}
z <- 1:20
for(i in 1:length(z)) {
  if(i %% 2==1) next
  print(i)
}
```

# Using != will skip if division DOES NOT EQUAL 0
```{r}
z <- 1:20
for(i in 1:length(z)) {
  if(i %% 2!=0) next
  print(i)
}
``` 
# Do we need a loop for this in R?
# NO
# Returning all odd elements with no loop
```{r}
z <- 1:20
zsub <- z[z %% 2 != 0] # Give me elements of z that are not divisible by 2
length(zsub)
zsub

z <- 1:20
zsub <- z[z %% 2 == 0] # Give me elements of z that are divisible by 2
length(zsub)
zsub
```

# TIP: Always look for vectorizing shortcuts to avoid loops if possible

#### But now, back to loops :D
```{r}
for (i in seq_along(zsub)) {
  cat("i =", i,"zub[i] " ,zsub[i], "\n")
}
```

# Creating a Function with for loops
Using the break function now in for loops to stop loops \n
Random walk example \n
NWKM \n
2/20/2025 \n

FUNCTION: ran_walk  \n
stochastic random walk \n
input: times = number of time steps  \n
       n1 = initial population size (= n[1])  \n
       lambda = finite rate of increase  \n
       noise_sd = sd of a normal distribution with mean 0  \n
output: vector n with population size > 0  \n
       until extinction, then NA  \n

if lamda = 1.0, there will be no change in time step  \n
if lamda = 1.1, there will be a 10% increase each time step  \n
if lamda = 0.98, there will be a 2% decrease each time step  \n

if lamda is > 1, we have a living organism  \n

with stochastic, each time step can be up or down, but on average  \n
will be 0 due to normal distribution  \n
will keep track of what happens until we hit 0  \n
in terms of real population, thats extinction <- this is where break comes into play  \n
 ------------------------------------------------------------------------------------
```{r}

library(ggplot2)
ran_walk <- function(times=100,
                     n1=50,
                     lamda = 1.00,
                     noise_sd = 10) {
  n <- rep(NA, times) # create output vector
  n[1] <- n1 # initial population size
  noise <- rnorm(n = times, mean = 0, sd = noise_sd) # create noise vector
  # on average, noise should be 0 but some will be pos and some will be neg
  for(i in 1:(times-1)) {
    n[i+1] <- lamda*n[i] + noise[i]
   if(n[i + 1] <= 0) {
      n[i + 1] <- NA
      cat("Population extinction at time", i+1, "\n")
      break
    }
  }
  return(n)
}

# Testing ran_walk() function
# This function is an example of something that needs to be done in a for-loop
# do we want stop instead of break?
ran_walk()
```

#### Using qplot with our new function <- but it was deprecated in 3.4.0 so be careful
```{r}
qplot(x = 1:100, y=ran_walk(), geom="line")
# The NAs help us keep x-axis at 100
```

#### Now, time for test cases - o boy
Checking our math, we see setting noise to 0 we get no change
i.e. population stays stable
```{r}
qplot(x = 1:100, y=ran_walk(noise_sd=0), geom="line")
```

#### Now testing  exponential growth
```{r}
qplot(x = 1:100, y=ran_walk(noise_sd = 0, lamda = 1.1), geom="line")
```

#### Now that we have seen exponential growth, lets see decay
```{r}
qplot(x = 1:100, y=ran_walk(noise_sd = 0, lamda = 0.98), geom="line")
```

#### Pure white noise
```{r}
qplot(x = 1:100, y=rnorm(n = 100, mean = 50), geom="line")
```

##### Could use round() function to keep integers
##### for sexual reproduction, what if all same sex? use unisexExtinct == runif(1) <= 2*(0.5)^n
##### environmental noise not random walk -- add rnorm(0,1)
##### measurement error -- add rnorm(0,1) for 0's w/o extinction

# Using double for loop - loop de loop
#### for test matrices, want them modularly sized and avoid square matrices if possible
#### So first, looping over rows
```{r}
m <- matrix(round(runif(20), digits = 2), nrow =5)
```

#### looping over rows
```{r}
for (i in 1:nrow(m)) { #could use for (i in seq_len(nrow(m)))
  m[i,] <- m[i,] + 1
}
print(m)
```

#### now looping over columns <- switched i with j, this will help with syntax and code reading
```{r}
for (j in 1:ncol(m)) { #could use for (i in seq_len(ncol(m)))
  m[,j] <- m[,j] + 1
}
print(m)
```

#### Now building a double for-loop, or nested for-loop
```{r}
m <- matrix(round(runif(20), digits = 2), nrow =5)
for (i in 1:nrow(m)) {
  for (k in 1:ncol(m)) {
    m[i,j] <- m[i,j] + i + j # taking number and adding values to it

  } # end of j loop
} # end of i loop
print(m)
```

## Creating function to sweep over parameters, including good formatting for function in code/comments <- from ecology
```{r}
# Function: Species area curve <- S = cA^z
# Creates power function relationship for S and A
# input:  A is a vector of island areas
#         c is the intercept constant
#         z is the slope constant
# output: S is a vector of species richness values
#
#
# ----------------------------------------------------------

species_area_curve <- function(A=1:5000, c = 0.5, z = 0.26) {
  S <- c*(A^z)
  return(S)
}
```

#### Using head function, returns first 6 elements
```{r}
head(species_area_curve())
```

## Creating Second function for plotting
```{r}
# function: species_area_plot
# plot species area curves with parameter values
# input: A = vector of areas
#        c = single value for c parameter
#        z = single value for z paremeter
#
# output: smoothed curve with parameters in graph
# -----------------------------------------------
# Also examples of using functions within functions

species_area_plot <- function(A=1:5000, c = 0.5, z = 0.26) {
  plot(x = A, species_area_curve(A,c,z),
       type = "l",
       xlab = "Island Area",
       ylab = "S",
       ylim=c(0,2500))
  mtext(paste("c =", c, "z = ", z), cex = 0.7) # This code will put z and c values on top of graphs when called
  # return()

}
species_area_plot()
```

#### Declaring global vairables <- often called at beginning of code
```{r}
c_pars <- c(100, 150, 175)
z_pars <- c(0.10, 0.16, 0.26, 0.3)
```

#### mfrow example
```{r}
par(mfrow=c(3,4)) # mfrow calls to graphics display, unique to base R graphics
# This will be in a matrix structure and uses our previous function
for (i in seq_along(c_pars)) { # will jump over c_pars values
  for (j in seq_along(z_pars)) { # will jump over z_pars values
    species_area_plot(c=c_pars[i], z=z_pars[j]) # should in total make 12 passes through this and get a grid of 12 plots
  }
}
```

# Comparing For and While loops
## looping with for
#### This is a good example of calculating something of unspecified length
#### break will stop when end point is reached
```{r}
cut_point <- 0.1
z <- NA
ran_data <- runif(100)
for (i in seq_along(ran_data)) {
  z <- ran_data[i]
  if (z < cut_point) break
}
print(z)
print(i)
```

## Similar to other languages, we will do a while loop
#### looping with while
#### in this example, we need to create a counter number, in this case cycle_number
```{r}
cut_point <- 0.1
z <- NA
cycle_number <- 0
```

## NOTE: 
if z = NA OR z is greater than or equal to cut_point, which is 0.1
loop will continue to run \n
loop will stop when z is less than the cut_point or is NA \n
instead of printing out loop number, since we do not have one, we print out cycle_number \n
Though, while loops are generally bad practice - for loops are generally better form \n
Worst part of while loop - unsure if it will ever end - I.E. it could never meet criteria then go on forever - no good \n
for example, if cut_point is something so low like 0.00001, it could run for a while,
no pun intended \n

```{r}
while (is.na(z) | z >= cut_point) {
  z <- runif(1)
  cycle_number <- cycle_number + 1

}
print(z)
print(cycle_number)
```

## Now using the repeat function
## looping with repeat
```{r}
z <- NA
cycle_number <- 0

repeat {
  z <- runif(1)
  cycle_number <- cycle_number + 1

  if (z <= cut_point) break
}
print(z)
print(cycle_number)
```

# Expand grid function <- Mucho importante
This function automatically sets up variables and pairs up unique combination of the two inputted values \n
c_pars is 3, z_pars is 4, so it gives us a data frame of 12 unique combinations \n
using c_pars and z_pars from earlier, lines 206 and 207
```{r}
expand.grid(c_pars,z_pars)
```

#### expand.grid examples for creating data sets
```{r}
df <- expand.grid(x = c_pars, y = z_pars) # x can be replaced with column/variable name for first value, y can be replaced for second value and so on
# can be down with more than 2 values
df2 <- expand.grid(x = c_pars, y = z_pars, z = c_pars*2) # now 36 values, 3*4*3 = 36
df2
head(df) # Returns first 6 rows
str(df) # You can see, the structure of df is a data.frame
df
```

## Creating sa_output function
#### This function gives us a 2 value list with x = max - min, and y = coefficient of variation
```{r}
##################################################
# function: sa_output
# Summary stats for species-area power function
# input: vector of predicted species richness (right now random numeric value)
# output: list of max-min called sum_stats, coefficient of variation
#-------------------------------------------------

sa_output <- function(S=runif(10)) {

  sum_stats <- list(s_gain=max(S)-min(S),s_cv=sd(S)/mean(S)) # list of  first: max value of s - min of s,
  # then taking the standard deviation of s divided by the mean of s for second list value

  return(sum_stats)
}
sa_output() # list is 2 values
```

# Build program body with a single loop through the parameters in modelFrame
#### Global variables, area vector, c_pars again, and z_pars
```{r}
area <- 1:5000
c_pars <- c(100,150,175)
z_pars <- c(0.10, 0.16, 0.26, 0.3)
```

#### set up model/data frame using expand.grid
```{r}
model_frame <- expand.grid(c=c_pars,z=z_pars)
model_frame$s_gain <- NA # adding column called s_gain with values NA
model_frame$s_cv <- NA # adding column called s_cv with values NA
print(model_frame)
```

#### cycle through model calculations - using multiple functions
#### starting from 1 to number of rows from model_frame - in our case, 12 rows
#### Replacing the values of NA from model_frame with our calculated values from our created functions
#### Notice, by using our functions and expand.grid we made this one_dimensional looping keeps values associated with appropriate rows
```{r}
for (i in 1:nrow(model_frame)) {

  # generate S vector
  # calling upon species_area_curve from earlier and creating a temp value
  # Using area vector from earlier which is 1 to 5000
  # model_frame[i,1] will take row i, col 1, which is our c_par, c
  # model_frame[i,2] will take row i, col 2, which is our z_par, z
  temp1 <- species_area_curve(A=area,
                              c=model_frame[i,1],
                              z=model_frame[i,2])
  # calculate output stats using our first temp value, creating a second temp value
  # temp2 <- sa_output(temp1) < could use temp2 but uneccessary
  # model_frame[i,c(3,4)] <- temp2
  # temp 2 is a list, and then were passing this list into a model_frame which we need   # to do since were passing in multiple columns into a data frame
  # pass results to columns in data frame
  model_frame[i,c(3,4)] <- sa_output(temp1)

}
print(model_frame)
```

## Parameter Sweeping Redux using ggplot to graph the data

#### First, global variables for loops for data
```{r}
area <- 1:5 #keep this very small initially
c_pars <- c(100,150,175)
z_pars <- c(0.10, 0.16, 0.26, 0.3)
```

#### set up model frame, this time with col S that has values NA and we have col A
#### model_frame will be 5 * 3 * 4 = 60 rows/combinations large
```{r}
model_frame <- expand.grid(c=c_pars,z=z_pars,A=area)
model_frame$S <- NA
model_frame
```

## 2 dimensional for loop through the parameters and fill column S with SA function values
```{r}
for (i in 1:length(c_pars)) {
  for (j in 1:length(z_pars)) {
    model_frame[model_frame$c==c_pars[i] & model_frame$z==z_pars[j],"S"] <-   species_area_curve(A=area,c=c_pars[i],z=z_pars[j])
  }
}
model_frame
```

## 1 dimensional for loop that assigns values of species_area_curve function to col S of model_frame, where the parameters for the function are the corresponding rows should be same values
#### Real power of expand.grid <- goes through 3 columns for 1 row in one call
#### Does not have to loop over parameters
```{r}
for (i in 1:nrow(model_frame)) {
  model_frame[i,"S"] <- species_area_curve(A=model_frame$A[i],
                                           c=model_frame$c[i],
                                           z=model_frame$z[i])
}
print(model_frame) # check by printing a data frame with limited parameter values
```

# Now graphing this with ggplot, first - call upon ggplot2 - notice we dont need for-loops for this <- Abusing R's vectorization
```{r}
library(ggplot2)
```

#### assigning plot of data of model_frame to p1 then adding comparison/graph details using geom_line
#### returns 12 graphs in one image
#### geom_line <- geometric/image ggplot will make, in this case a line with:
#### x axis is area (1,2,3,4,5), y value/axis is s (what we calculated)
#### using facet_grid, our graphs are set up to show graphs for each c and z combination
#### Therefore, 12 graphs
#### don't need $ for referencing data.col names as we specified data in our p1 variable
```{r}
p1 <- ggplot(data=model_frame)
p1 + geom_line(mapping= aes(x=A,y=S)) +
  facet_grid(c~z)
```

#### returns 3 graphs, each graph has 4 lines (values from z) for each c, giving us 3 graphs
#### Grouped by c values
#### notice the change in facet_grid, now it is saying all z values for each c
```{r}
p2 <- p1
p2 + geom_line(mapping=aes(x=A,y=S,group=z)) +
  facet_grid(.~c)
```

#### returns 4 graphs, each graph has 3 lines (values from c) for each z, giving us 4 graphs
#### Grouped by z values
#### notice the change in facet_grid, now it is saying all c values for each z
```{r}
p3 <- p1
p3 + geom_line(mapping=aes(x=A,y=S,group=c)) +
  facet_grid(z~.)
```
